On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-26 Manuscript/manuscript.tex
Update after tuesdays meeting
finished first pass of manuscript

%1)
Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 in his preliminary heat and 0.101 in his semifinal.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championship 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.

%\eds{final two sentence seems out of place here; consider moving elsewhere?}
%\of{Moved to the end of paragraph 4 of introduction}

@@

\subsection{World Championship}

@@

Data was taken from the World Athletics 
\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles 
and the women's 100 meter hurdles from 1999 to 2022.  The dependent
variable interest is reaction time, and the independent variables are year, 
stage of competition, total time, and gender.  \eds{see earlier comment - 
are you using any of these as covariates?}
The data can be found in the appendix~\ref{sec:Appendix}. Factors such as name, 
country, and identification number were not considered important for the 
purposes of this research besides the times for Devon Allen and the other 
United States athletes mentioned above. \eds{not sure which other US athletes 
you are referring to in the previous sentence} 
The variable ``Stage" refers to whether the observation occurred during a 
%heat (preliminary round)  
preliminary heat denoted by ``H", semifinal heat (of which there are either two
or three each year) denoted by ``S", or final heat denoted by ``F". The variable 
``Gender" has levels ``M" and ``F" for ``male" and ``female" respectively.  The
variable ``Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly). There is no statistical meaning to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary heat of 2022 and Batch
111 corresponds to the 1999 finals). %Included are
%box plots of the reaction times of the men's finals and the men's semifinals and
%finals pooled together.  
For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
%We will refer to the latter as the ``pooled data"
throughout the paper.  %The heats or preliminary rounds were not studied because
The preliminary heats were not studied because athletes react faster as the 
competition progresses \citep[e.g.,][]{zhang2021correlation}. % Athletes react 
%faster in the finals than the semifinals which are faster than the heats 
%\citep{zhang2021correlation}.  
\eds{if you are really not using ``H'' data at all, it does not appear necessary
to spend time defining earlier.  You could just say directly when you define 
stage that you only consider semifinals and finals and cite the Zhang paper for 
rationale}
If we want to determine if Allen's reaction time
was reasonable, then we thus need to look at the men's final data, with the men's
pooled data being included in order to increase the sample size.  
\eds{previous sentence is not clear; pooled data already includes final data}
In 2022, there
were only five data points due to two disqualifications and one athlete who 
qualified but did not compete.
The data under consideration is best visualized using side-by-side box plots, as
shown in Figure~\ref{fig:Boxplots}.  It is clear that the 2022 box plot for both 
the men's finals and semifinals data is lower (indicating faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\citet{brosnan2017effects} found reaction times of 0.156 seconds \eds{mean?  median?} for men when looking at data from 1999 to 2014.

@@

\subsection{Beyond World Championship}
The motivation for looking beyond World Championship data came from an 
initial qualitative analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.
\eds{be consistent with how you refer to the US championship:
``United States Track and Field Championships" and ``USA Track and Field 
Championships" are both being used; please pick one and use that throughout}
%This section was motivated from a qualitative analysis that started by looking
%at how United States athletes reacted at the 2022 United States Track and Field
%Championships. 
From June 23-26 2022, the United States held its Track
and Field Championship at Hayward 
Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. %the exact same venue as used 
%for the World Track and Field Championship a month later.  
%Thus, a baseline is able to be established for the four USA athletes who 
Thus, we can establish a baseline for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates. \eds{statistically significantly faster? If you used a statistical 
test, it should not be described in this section.  If you didn't use a 
statistical test, you should replace the word `significantly' with something 
else (substantially?)} 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. 
%``USA" denotes
%the USA Track and Field Championships, ``World" denotes the World Track and Field
%Championships. All numbers listed in the table are the reaction
%time for the respective athletes in seconds.

@@

This dataset is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).

@@

The World Championship data and the data beyond the World Championship were
analysed with GLMM and rank-based approaches, respectively.
\eds{need consistency with ``Championship" or ``Championships - I am not sure 
which one is correct}

\subsection{GLMM}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championship, the final model that we chose was a gamma
mixed-effect model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generlized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. The link function $g$ can be the logarithm function to
ensure the positivity of the conditional mean.
The gamma mixed-effect model can be fit with the \texttt{glmer()} function in R
package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $v = ab^2$. The
variance as a function of the mean $\mu$ in the GLM sense is
$v(\mu) = \mu^2 / a$, which implies that the dispersion parameter
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ij}$ given the venue effect $z_i$
has shape $a = 1 / \phi$ and scale $b = g^{-1}(\alpha + z_i) \phi$. The marginal
distribution of $Y_{ij}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. A large number of
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We want the probability of a reaction time being less than 0.1
seconds in order to gauge if that is a reasonable disqualification barrier.

As for the random effect, it was possible to include as well a race effect.
From our exploration, however, the race effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the race reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the race
effect, but are all factors that could potentially impact the venue effect.

@@

To test the conjecture that the 2022 World Champion timing device may have led
to faster recorded reaction times, we compare the reaction times of the same
athletes who have attended both the 2022 World Champion and other competitions.
In this setting, we have clustered data with subunit grouping. In particular,
each athlete is a cluster and the multiple reaction times from the same athlete
can be from either the 2022 World Championship or otherwise.
Let $X_{ij}$ be the $j$th reaction time of athelete~$i$, $i = 1, \ldots, n$,
$j = 1, \ldots, m_i$ where $m_i$ is the number of observations from
athelete~$i$. Let $\delta_{ij}$ be the group indicator of $X_{ij}$; $\delta_{ij}
= 1$ if $X_{IJ}$ is in group~1 and $\delta_{ij} = 0$ otherwise. Atheletes are
assumed to be independent, while subunit observations from the same athelete are
not. The null hypothesis $H_0$ to be tested is that there is no difference
between the two groups; i.e., the distribution of $X_{ij}$ remains the same
regardless of the group indicator $\delta_{ij}$.

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the psudo-sample.
The test statistic $S$ is the average of $W^*$ averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a statistic $Z$ which follows standard normal distribution
asymptotically \citep[p.910]{datta2005rank}. An implementation of this method is
available from the \texttt{clusWilcox.test()} function with
\texttt{method = ``ds'} from R package \texttt{clusrank}
\citep{jiang2017wilcoxon}.
%% In this case, athletes who have more
%% data are generally thought to be faster, as they have made it further in the
%% competition.  Clusters range from two to six, with two indicating that at each
%% competition they only made it to the heats, but a cluster size of six indicates
%% that they made the finals in both competitions.  While the number of times
%% athletes compete is based off their overall time not their reaction time, the
%% two are correlated \citep{delalija2008reaction}.
%% Group is either designated to be 1 or 2, with 1 corresponding to either the
%% national level data or the previous world championship data and 2 corresponding
%% to the 2022 data.

@@

\of{Add something like this to the start of the paragraph?}
We may want to consider using all four data sets for this problem because
comparing all the data to the data excluding 2022 can give us a qualitative
assessment of how impactful 2022 was.

@@

\jy{Close captions with a period.}
\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets.}
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Older Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Older Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

At a size of $n=10,000,000$, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:
\jy{label the table and reference to it; clean it up}
\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than 0.1, 0.09, and
  0.08 for various data sets.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & P(RxnTime $<$ 0.1) & P(RxnTime $<$ 0.09) & P(RxnTime $<$ 0.08) \\ 
   \midrule
   all finals & $8.56\cdot10^{-4}$ & $4.1\cdot10^{-5}$ & $6.0\cdot10^{-7}$ \\ 
   old finals & $4.39\cdot10^{-4}$ & $1.63\cdot10^{-5}$ & $2.0\cdot10^{-5}$ \\
   all pooled & $6.86\cdot10^{-3}$ & $1.21\cdot10^{-3}$ & $1.3\cdot10^{-4}$ \\
   old pooled & $5.76\cdot10^{-3}$ & $9.93\cdot10^{-4}$ & $1.0\cdot10^{-4}$\\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

Using the gamma function with link set to log, the probability of the function being
less than 0.1 seconds can be calculated to see if Allen's disqualification was an anomaly.
Prior to the 2022 World Championship, there was a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals as shown in table~\ref{tab:Sim_probability}.
That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (The World Championship is biennial), every 569 years there
should be a time as extreme as Allen's.  This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

It is interesting that the probability of observing a reaction time below 0.1
seconds increases drastically for the finals data and increases by a smaller margin
for the pooled data set after 2022 is included in the data set.  By including 2022
in the data, the chances of observing an extreme time are increased, showing that
2022 is a high outlier for reaction times compared to other years.

Table ~\ref{tab:Sim_probability} shows that changing the reaction time barrier from 0.1 seconds to
0.08 seconds drastically reduces the chances of observing a reaction time that
would break the barrier.  For the dataset of all finals data, the probability
associated with observing a reaction time below 0.08 seconds was $6\cdot10^{-7}$.
If IAAF were to change the reaction time the way that \citet{komi2009iaaf}'s paper
suggests, the probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.

@@

When the clusrank rank sum test was performed, the ``ds" method produced a z score of
2.9751 and p value of 0.001464.  This value is for a one-sided test and show that
the mean reaction time on average in 2019 was larger than the mean reaction time
in 2022 for athletes who competed in both championships at an alpha level of 0.01.
This provides good evidence that athletes who competed in both 2019 and 2022 got
faster during that timespan.

Repeating the data collection steps and
again performing a clustered analysis for a Wilcoxon ranked sum test
results in two new z scores and p values.  The ``ds" method produced a z score of
3.6069 and p value of 0.000155.  This result is highly significant and shows that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships even for an alpha level of 0.001.

>>>>

%1)
Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 in his preliminary heat and 0.101 in his semifinal.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championship 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.
\eds{you mention in the final section that the next world championship is in 
2023, so he won't need to wait two years if that is true}
%\eds{final two sentence seems out of place here; consider moving elsewhere?}
%\of{Moved to the end of paragraph 4 of introduction}

@@

\subsection{World Championship}\label{sec:dataworld}

@@

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles 
and the women's 100 meter hurdles from 1999 to 2022.  \eds{see earlier comment - 
are you using any of these as covariates?}
The data source can be found in the Appendix.%~\ref{sec:Appendix}. 
Factors such as name, 
country, and identification number were not considered important for the 
purposes of this research besides the times for Devon Allen and the other 
United States athletes mentioned above. \eds{not sure which other US athletes 
you are referring to in the previous sentence} 
The variable ``Stage" refers to whether the observation occurred during a 
%heat (preliminary round)  
preliminary heat denoted by ``H", semifinal heat (of which there are either two
or three each year) denoted by ``S", or final heat denoted by ``F". The variable 
``Gender" has levels ``M" and ``F" for ``male" and ``female" respectively.  The
variable ``Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly). There is no statistical meaning to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary heat of 2022 and Batch
111 corresponds to the 1999 finals). %Included are
%box plots of the reaction times of the men's finals and the men's semifinals and
%finals pooled together.  
For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
%We will refer to the latter as the ``pooled data"
throughout the paper.  %The heats or preliminary rounds were not studied because
The preliminary heats were not studied because athletes react faster as the 
competition progresses \citep[e.g.,][]{zhang2021correlation}. % Athletes react 
%faster in the finals than the semifinals which are faster than the heats 
%\citep{zhang2021correlation}.  
\eds{if you are really not using ``H'' data at all, it does not appear necessary
to spend time defining earlier.  You could just say directly when you define 
stage that you only consider semifinals and finals and cite the Zhang paper for 
rationale}
If we want to determine if Allen's reaction time
was reasonable, then we thus need to look at the men's final data, with the men's
pooled data being included in order to increase the sample size.  
\eds{previous sentence is not clear; pooled data already includes final data}
In 2022, there
were only five data points due to two disqualifications and one athlete who 
qualified but did not compete.
The data under consideration is best visualized using side-by-side box plots, as
shown in Figure~\ref{fig:Boxplots}.  It is clear that the 2022 box plot for both 
the men's finals and semifinals data is lower (indicating faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\citet{brosnan2017effects} found reaction times of 0.156 seconds \eds{mean?  median?} for men when looking at data from 1999 to 2014.

@@

\subsection{Beyond World Championship}\label{sec:databeyond}
The motivation for looking beyond World Championship data came from an 
initial qualitative \eds{exploratory instead of qualitative?}
analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.
\eds{be consistent with how you refer to the US championship:
``United States Track and Field Championships" and ``USA Track and Field 
Championships" are both being used; please pick one and use that throughout}
%This section was motivated from a qualitative analysis that started by looking
%at how United States athletes reacted at the 2022 United States Track and Field
%Championships. 
From June 23-26 2022, the United States held its Track
and Field Championship at Hayward 
Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. %the exact same venue as used 
%for the World Track and Field Championship a month later.  
%Thus, a baseline is able to be established for the four USA athletes who 
Thus, we can establish a baseline for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teammates. \eds{statistically significantly faster? If you used a statistical 
test, it should not be described in this section.  If you didn't use a 
statistical test, you should replace the word `significantly' with something 
else (substantially?)} 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. 
%``USA" denotes
%the USA Track and Field Championships, ``World" denotes the World Track and Field
%Championships. All numbers listed in the table are the reaction
%time for the respective athletes in seconds.

@@

This dataset is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).  \eds{this is where you can put the relevant 
information about the other competitions that was commented out in the 
introduction.}
\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}

@@

The %World Championship data and the data eyond the World Championship 
data described in Sections~\ref{sec:dataworld} and~\ref{sec:databeyond} were
analysed with GLMM and rank-based methods, respectively.
\eds{need consistency with ``Championship" or ``Championships - I am not sure 
which one is correct}

\subsection{GLMM}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championship, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
\eds{we know that many of the athletes compete in multiple world championships,
how is this athlete clustering being accounted for here?}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. The link function $g$ can be the logarithm function to
ensure the positivity of the conditional mean.
\eds{Did you try an identity link also? Lo and Andrews (2015) suggest that the 
Gamma family with an identity link works well for their reaction-time data.
It doesn't seem like a good idea in general because it is likely that some 
negative predicted responses will occur either during model fitting or in the 
final model, but the internet seems to favor this link so a reviewer may ask 
about it too.}
The gamma mixed-effect model can be fit with the \texttt{glmer()} function in R
package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $v = ab^2$. The
variance as a function of the mean $\mu$ in the GLM sense is
$v(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ij}$ given the venue effect $z_i$
has shape $a = 1 / \phi$ and scale $b = g^{-1}(\alpha + z_i) \phi$. The marginal
distribution of $Y_{ij}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. A large number of
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
 being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

As for the random effect, it was possible to include a race 
\eds{heat? when I first read this, I thought you meant race as in ethnicity} 
effect, as well.
From our exploration, however, the race effect is very difficult to explain or 
interpret as it means that some heats of races were either
faster or slower and that everyone in the race reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier, however, to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art. Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the race
effect, but are all factors that could potentially impact the venue effect.

@@

To test the conjecture that the 2022 World Champion timing device may have led
to faster recorded reaction times, we compare the reaction times of the same
athletes who have attended both the 2022 World Champion and other competitions.
In this setting, we have clustered data with subunit grouping. In particular,
each athlete is a cluster and the multiple reaction times from the same athlete
can be from either the 2022 World Championship or otherwise.
Let $X_{ij}$ be the $j$th reaction time of athlete~$i$, $i = 1, \ldots, n$,
$j = 1, \ldots, m_i$ where $m_i$ is the number of observations from
athelete~$i$. Let $\delta_{ij}$ be the group indicator of $X_{ij}$; $\delta_{ij}
= 1$ if $X_{ij}$ is in group~1 (2022 World Championship) and $\delta_{ij} = 0$ 
otherwise. Athletes are
assumed to be independent, while subunit observations from the same athlete are
not. The null hypothesis $H_0$ to be tested is that there is no difference
between the two groups; i.e., the distribution of $X_{ij}$ remains the same
regardless of the group indicator $\delta_{ij}$.

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the pseudo-sample.
The test statistic $S$ is the average of $W^*$ averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a statistic $Z$ which follows standard normal distribution
asymptotically \citep[p.910]{datta2005rank}. An implementation of this method is
available from the \texttt{clusWilcox.test()} function with
\texttt{method = ``ds'} from R package \texttt{clusrank}
\citep{jiang2017wilcoxon}.
\eds{are you including preliminary heat data as possible within cluster 
observations in this analysis?  Based on your 
discussion in Section 2.1 about how they are not comparable, it may not be fair 
to include them here?  Or perhaps do the analysis with and without prelim heat
data to see if it makes a difference?}

%% In this case, athletes who have more
%% data are generally thought to be faster, as they have made it further in the
%% competition.  Clusters range from two to six, with two indicating that at each
%% competition they only made it to the heats, but a cluster size of six indicates
%% that they made the finals in both competitions.  While the number of times
%% athletes compete is based off their overall time not their reaction time, the
%% two are correlated \citep{delalija2008reaction}.
%% Group is either designated to be 1 or 2, with 1 corresponding to either the
%% national level data or the previous world championship data and 2 corresponding
%% to the 2022 data.

@@

\of{Add something like this to the start of the paragraph?}
We may want to consider using all four data sets for this problem because
comparing all the data to the data excluding 2022 can give us a qualitative
assessment of how impactful 2022 was.
\eds{Agree with JY, but you need to describe what these four datasets are.}

Table~\ref{tab:Gamma_parameters} summarizes the results from the gamma model 
from different data sets.

@@

\jy{Close captions with a period.}
\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets. 
	\eds{Int=intercept?}}
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Older Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Older Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

\eds{be consistent with the order of the rows between 
Tables~\ref{tab:Gamma_parameters} and~\ref{tab:Sim_probability}}

\jy{label the table and reference to it; clean it up}
\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than 0.1, 0.09, and
  0.08 for various data sets.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & P(RxnTime $<$ 0.1) & P(RxnTime $<$ 0.09) & P(RxnTime $<$ 0.08) \\ 
   \midrule
   all finals & $8.56\cdot10^{-4}$ & $4.1\cdot10^{-5}$ & $6.0\cdot10^{-7}$ \\ 
   old finals & $4.39\cdot10^{-4}$ & $1.63\cdot10^{-5}$ & $2.0\cdot10^{-5}$ \\
   all pooled & $6.86\cdot10^{-3}$ & $1.21\cdot10^{-3}$ & $1.3\cdot10^{-4}$ \\
   old pooled & $5.76\cdot10^{-3}$ & $9.93\cdot10^{-4}$ & $1.0\cdot10^{-4}$\\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

Using the gamma function with link set to log, the probability of the reaction 
time being less than 0.1 seconds can be calculated to see if Allen's 
disqualification was an anomaly. \eds{explain how simulation was used here.}
At a size of $n=10,000,000$ \eds{use different letter as $n$ was already used 
for number of athletes, $B?$}, the calculated probabilities of a reaction time 
being below 0.1 seconds for the various different data sets are shown in 
Table~\ref{tab:Sim_probability}. Prior to the 2022 World Championship, there was 
a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals.
That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (the World Championship is biennial), every 569 years there
should be a time as extreme as Allen's. This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

When the 2022 data is included in the analysis, it is interesting that the 
probability of observing a reaction time below 0.1 seconds increases drastically 
for the finals data and increases by a smaller margin for the pooled data.  By
including 2022 in the data, the chances of observing an extreme time are 
increased, suggesting that 2022 is a high outlier for reaction times compared to
 other years.

Table ~\ref{tab:Sim_probability} also shows that changing the reaction time 
barrier from 0.1 seconds to 0.08 seconds drastically reduces the chances of 
observing a reaction time that would break the barrier.  For the dataset of all 
finals data, the probability
associated with observing a reaction time below 0.08 seconds was $6\cdot10^{-7}$.
If IAAF \eds{or World Athletics?} were to change the reaction time the way that 
\citet{komi2009iaaf}
suggests, the probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.

@@

When the clusrank rank sum test was performed, the ``ds" method produced a z score of
2.9751 and p value of 0.001464.  \eds{what is the ds method?} 
This value is for a one-sided test and show that
the mean reaction time on average in 2019 was larger (slower) than the mean 
reaction time
in 2022 for athletes who competed in both championships at an alpha level of 0.01.
This provides good evidence that athletes who competed in both 2019 and 2022 got
faster during that timespan.  \eds{That is not the only explanation}

Repeating the data collection steps and
again performing a clustered analysis for a Wilcoxon ranked sum test
results in two new z scores and p values. 
\eds{I am not sure why you are repeating the data collection steps and
again performing a clustered analysis here.  Also, if there are two new z-scores
and p values, why are you only reporting one?}
  The ``ds" method produced a z score of
3.6069 and p value of 0.000155. This result is highly significant and shows that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships even for an alpha level of 0.001.
